---
title: "Project 2 – TMA4300"
author: "Karine Grande & Thea Boge"
date: "2026-21-03"
output: pdf_document
header-includes:
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{multirow}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

```{r, echo=FALSE}
library(glmmTMB)
library(coda)    
library(RTMB)
library(kableExtra)
library(mvtnorm)
```

# Problem 1
We suppose that the joint density of $U$, $Z$ up to a normalizing constant is 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]

## a) 
If a joint density is known up to proportionality, the full conditional of one variable is obtained by 
\[
f_{U|Z}(u|z) \propto f^*_{U,Z}(u,z) \quad \text{as a function of } u
\]

### Conditional ditribution of $U|Z=z$

Fix $z$. The joint density becomes 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]
As a function of $u$, this is constant on the interval $(0,e^{-z^2/2})$. 

The normalizing constant is
\[
\int_0^{e^{-z^2/2}} 1 \, du =e^{-z^2/2}.
\]

Therefore,
\[
f_{U \mid Z}(u \mid z) = \frac{1}{e^{-z^2/2}} = e^{z^2/2},\quad 0 < u < e^{-z^2/2}.
\]

Hence,
\[
U \mid Z = z \sim \mathrm{Unif} \big(0, e^{-z^2/2}\big).
\]

### Conditional ditribution of $Z|U=u$
Fix $u$. The support condition is $0 < u < e^{-z^2/2}$.

Solving for $z$:
\[
u < e^{-z^2/2} \quad \Longleftrightarrow \quad \log u < -\frac{z^2}{2}
\]

\[
-2 \log u > z^2
\]

\[
|z| < \sqrt{-2 \log u}
\]

Therefore, the support of the conditional distribution is
\[
z \in (-\sqrt{-2 \log u}, \sqrt{-2 \log u}).
\]

Within this interval the joint density is constant and equal to 1. The normalizing constant is therefore the length of the interval:
$$
\int_{-\sqrt{-2 \log u}}^{\sqrt{-2 \log u}} 1 \, dz = 2\sqrt{-2 \log u}.
$$

Thus, the full conditional density of \( Z \mid U = u \) is

\[
f_{Z \mid U}(z \mid u) = \frac{1}{2\sqrt{-2 \log u}},\quad |z| < \sqrt{-2 \log u},
\]

Thus,
\[
Z \mid U = u \sim \mathrm{Unif}\big(-\sqrt{-2 \log u}, \sqrt{-2 \log u}\big).
\]

### Gibbs Sampler

Given an initial valid point \( (U^{(0)}, Z^{(0)}) \) satisfying
\( 0 < U^{(0)} < e^{-(Z^{(0)})^2/2} \),
iterate for \( t = 1, \dots, n \):

\begin{enumerate}
    \item Sample
    \[
    U^{(t)} \sim 
    \mathrm{Unif}\left(0, e^{-(Z^{(t-1)})^2/2}\right).
    \]

    \item Define
    \[
    a^{(t)} = \sqrt{-2 \log U^{(t)}}.
    \]

    \item Sample
    \[
    Z^{(t)} \sim 
    \mathrm{Unif}\left(-a^{(t)},\, a^{(t)}\right).
    \]
\end{enumerate}

This iterative scheme defines a Markov chain whose stationary distribution is the joint density \( f_{U,Z}(u,z) \).The marginal stationary distribution of \( Z \) is \( \mathcal{N}(0,1) \).

```{r gibbs_sampler, echo=TRUE}
gibbs_normal <- function(n, u_init = 0.5, z_init = 0) {
  
  U <- numeric(n)
  Z <- numeric(n)
  
  U[1] <- u_init
  Z[1] <- z_init
  
  for (t in 2:n) {
    U[t] <- runif(1, 0, exp(-(Z[t-1]^2)/2))
    a <- sqrt(-2 * log(U[t]))
    Z[t] <- runif(1, -a, a)
  }
  Z
}
```

```{r 2, echo=FALSE}
set.seed(123)
Z <- gibbs_normal(10000)

Z_post <- Z[1001:10000]

mean(Z_post)
var(Z_post)

hist(Z_post, probability = TRUE, breaks = 40)
curve(dnorm(x), col="red", add=TRUE)
```
The posterior sample mean is approximately
\[
\hat{\mu} \approx -0.019,
\]
and the empirical variance is
\[
\hat{\sigma}^2 \approx 0.97.
\]

Both values are close to the theoretical mean and variance,
\[
\mathbb{E}[Z] = 0,\qquad \mathrm{Var}(Z) = 1,
\]
of a standard normal distribution.

The histogram aligns well with the standard normal density curve, confirming that the Gibbs sampler correctly targets
\[
Z \sim \mathcal{N}(0,1).
\]

## b) Marginal distribution of Z from the Gibbs sampler
We determine the marginal distribution of \( Z \) by integrating out \( U \).

The marginal density of \( Z \) is

\[
f_Z(z)
=
\int_0^{e^{-z^2/2}} 1 \, du
=
e^{-z^2/2}.
\]

Thus,

\[
f_Z(z) \propto e^{-z^2/2}.
\]


Since
\[
\int_{-\infty}^{\infty} e^{-z^2/2} \, dz
=
\sqrt{2\pi},
\]
the normalized density is

\[
f_Z(z)
=
\frac{1}{\sqrt{2\pi}} e^{-z^2/2}.
\]

Therefore,
\[
Z \sim \mathcal{N}(0,1).
\]
Hence, discarding the \( U \)-values from the Gibbs sampler yields a sample from a standard normal distribution.

## c) Correlation between successive Gibbs samples

From the Gibbs sampler,

$$
Z_n \mid U_n \sim \mathrm{Unif}(-\sqrt{-2 \log U_n}, \sqrt{-2 \log U_n})
$$

which is symmetric, so

$$
\mathbb{E}[Z_n \mid U_n] = 0.
$$

In stationarity, $\mathbb{E}[Z_n] = \mathbb{E}[Z_{n-1}] = 0$, hence

$$
\mathrm{Cov}(Z_{n-1}, Z_n) = \mathbb{E}[Z_{n-1} Z_n].
$$

Using iterated expectation and conditional independence given $U_n$,

$$
\mathbb{E}[Z_{n-1} Z_n]
=
\mathbb{E}\bigl[
\mathbb{E}[Z_{n-1} \mid U_n]\,
\mathbb{E}[Z_n \mid U_n]
\bigr]
= 0.
$$

Thus,

$$
\mathrm{Cov}(Z_{n-1}, Z_n) = 0.
$$

So successive samples are uncorrelated, but not independent.

For $Z_n \mid U_n \sim \mathrm{Unif}(-a,a)$ with $a = \sqrt{-2\log U_n}$,

$$
\mathbb{E}[Z_n^2 \mid U_n] = \frac{a^2}{3} = \frac{-2\log U_n}{3}.
$$

Hence,

$$
\mathbb{E}[Z_{n-1}^2 \mid U_n] = \mathbb{E}[Z_n^2 \mid U_n].
$$

Using total covariance,

$$
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Var}\left(\frac{-2\log U_n}{3}\right).
$$

Let $T = -2\log U_n$. In stationarity, $T \sim \chi^2_3$, so $\mathrm{Var}(T)=6$. Therefore,

$$
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\frac{6}{9}
=
\frac{2}{3}.
$$

Thus, squared samples are positively correlated.



# Problem 2
## a) maximum likelihood estimates
We first obtain maximum likelihood estimates of the model parameters by fitting a random intercept model to the data.

```{r, echo=FALSE, }
load(url("https://www.math.ntnu.no/emner/TMA4300/2026v/random-intercept-data.RData"))
mod <- glmmTMB(y ~ (1|i), data=data)
summary(mod)
```

The maximum likelihood estimate of the overall mean is $\hat{\mu} = 10.1888$. The estimated variance of the random intercepts is $\widehat{\mathrm{Var}}(\gamma_i) = 0.2301$, corresponding to $1/\nu$, and the estimated residual variance is $\widehat{\mathrm{Var}}(\epsilon_{ij}) = 0.7560$, corresponding to $1/\tau$. These estimates will later be used to check that the Gibbs sampler implementation produces reasonable results.

## b) Full conditional distributions

From the hierarchical model

\[
y_{ij} \mid \mu,\gamma_i,\tau \sim N(\mu+\gamma_i,\tau^{-1}),
\]

\[
\gamma_i \mid \nu \sim N(0,\nu^{-1}),
\]

with priors

\[
\pi(\mu)\propto 1, \qquad \tau \sim \text{Gamma}(\alpha,\alpha), \qquad \nu \sim \text{Gamma}(\alpha,\alpha),
\]

we derive the following full conditional distributions.

Let \(N=\sum_{i=1}^{m} n_i\).


Full conditional of \( \mu \). Combining the likelihood with the flat prior for \(\mu\), we obtain

\[
\mu \mid \gamma,\tau,y \sim N(\bar{y}_\gamma,(N\tau)^{-1}),
\]

where

\[
\bar{y}_\gamma
=
\frac{1}{N}
\sum_{i=1}^{m}\sum_{j=1}^{n_i}(y_{ij}-\gamma_i).
\]

Full conditional of \( \gamma_i \). Using the likelihood contribution from cluster \(i\) together with the prior \(\gamma_i \sim N(0,\nu^{-1})\), we obtain

\[
\gamma_i \mid \mu,\tau,\nu,y
\sim
N(m_i,(\nu+n_i\tau)^{-1}),
\]

where

\[
m_i
=
\frac{\tau\sum_{j=1}^{n_i}(y_{ij}-\mu)}{\nu+n_i\tau}.
\]


Full conditional of \( \tau \). Let

\[
S=\sum_{i=1}^{m}\sum_{j=1}^{n_i}(y_{ij}-\mu-\gamma_i)^2.
\]

Combining the Gaussian likelihood with the prior \(\tau \sim \text{Gamma}(\alpha,\alpha)\) gives

\[
\tau \mid \mu,\gamma,\nu,y
\sim
\text{Gamma}\left(\alpha+\frac{N}{2},\,\alpha+\frac{S}{2}\right).
\]


Full conditional of \( \nu \). Using the prior for \(\gamma_i\) and \(\nu \sim \text{Gamma}(\alpha,\alpha)\), we obtain

\[
\nu \mid \mu,\gamma,\tau,y
\sim
\text{Gamma}\left(\alpha+\frac{m}{2},\,\alpha+\frac{1}{2}\sum_{i=1}^{m}\gamma_i^2\right).
\]

## c) Implementation of a Gibbs sampler

```{r, echo=TRUE}
gibbs_sampler <- function(data, n_iter = 20000, alpha = 0.0001){

  y <- data$y
  i <- as.numeric(factor(data$i))

  m <- length(unique(i))
  N <- length(y)


  n_i <- table(i)

  mu <- mean(y)
  gamma <- rep(0, m)
  tau <- 1
  nu <- 1


  samples <- matrix(NA, n_iter, m + 3)

  for (iter in 1:n_iter){

    y_gamma <- y - gamma[i]
    mu <- rnorm(1, mean(y_gamma), sqrt(1/(N * tau)))

    for (k in 1:m){

      y_k <- y[i == k]
      n_k <- length(y_k)

      var_gamma <- 1 / (nu + n_k * tau)
      mean_gamma <- (tau * sum(y_k - mu)) / (nu + n_k * tau)

      gamma[k] <- rnorm(1, mean_gamma, sqrt(var_gamma))
    }

    res <- y - mu - gamma[i]
    tau <- rgamma(1, alpha + N/2, alpha + sum(res^2)/2)

    nu <- rgamma(1, alpha + m/2, alpha + sum(gamma^2)/2)

    samples[iter, ] <- c(mu, gamma, 1/tau, 1/nu)
  }

  colnames(samples) <- c(
    "mu",
    paste0("gamma_", 1:m),
    "sigma2_eps",
    "sigma2_gamma"
  )

  samples <- mcmc(samples)

  return(samples)
}
```

```{r, echo=FALSE}
samples <- gibbs_sampler(data, 20000)
colMeans(samples)
```

The Gibbs sampler was run for 20000 iterations to sample from the joint posterior distribution of the model parameters. The posterior mean of the global intercept is approximately $\mu \approx 10.19$, which is very close to the maximum likelihood estimate obtained in part (a). The posterior mean of the residual variance is approximately $1/\tau \approx 0.80$, and the posterior mean of the random intercept variance is approximately $1/\nu \approx 0.25$. These values are close to the corresponding maximum likelihood estimates from part (a), namely $0.756$ and $0.2301$. This suggests that the Gibbs sampler is producing reasonable posterior estimates and that the implementation appears to be correct.


## d) `summary`function
To evaluate the behaviour of the Gibbs sampler, we inspect trace plots for the main parameters and variance components.
```{r, echo=FALSE}
par(mfrow = c(3,1))
plot(samples[,1], type = "l", main = expression(mu), ylab = expression(mu), xlab = "Iteration")
plot(samples[,22], type = "l", main = expression(1/tau), ylab = expression(1/tau), xlab = "Iteration")
plot(samples[,23], type = "l", main = expression(1/nu), ylab = expression(1/nu), xlab = "Iteration")
plot(log(samples[,23]), type = "l", main = expression(log(1/nu)),
     ylab = expression(log(1/nu)), xlab = "Iteration")
```
The trace plots show that the Markov chains fluctuate around stable levels with no visible trends, suggesting that the Gibbs sampler mixes reasonably well. The posterior mean of the global intercept $\mu$ is around 10.2, which is very close to the maximum likelihood estimate obtained in part (a). The trace plot for $\mu$ shows tight fluctuations around this level and the posterior density appears approximately normal.

The residual variance $1/\tau$ fluctuates around approximately 0.8 and also appears stable across iterations. Its posterior distribution is slightly right-skewed, which is typical for variance parameters.

The random intercept variance $1/\nu$ behaves somewhat differently. Its trace plot shows larger variability and occasional spikes toward larger values. The posterior density is strongly right-skewed, with most of its mass concentrated near small values. This indicates that there is greater uncertainty in the random intercept variance than in the residual variance.

Examining the trace plot of $\log(1/\nu)$ helps visualize this behavior more clearly. On the log scale the chain appears more regular, while still showing occasional excursions corresponding to large values of $1/\nu$. Overall, the trace plots suggest that the Gibbs sampler mixes reasonably well, although the random intercept variance is clearly the most variable parameter in the model.

## e) Prior distribution of the proportion of variance

From (6),
\[
p = \frac{\nu^{-1}}{\nu^{-1}+\tau^{-1}} = \frac{\tau}{\tau+\nu}.
\]

With independent priors
\[
\tau \sim \Gamma(\alpha,\alpha), \quad \nu \sim \Gamma(\alpha,\alpha),
\]
and the standard result that if
\[
X \sim \Gamma(a,\lambda), \quad Y \sim \Gamma(b,\lambda),
\]
then
\[
\frac{X}{X+Y} \sim \mathrm{Beta}(a,b),
\]
it follows that
\[
p \sim \mathrm{Beta}(\alpha,\alpha).
\]

With \(\alpha = 0.0001\),
\[
p \sim \mathrm{Beta}(0.0001, 0.0001).
\]

This prior is strongly U-shaped, placing most mass near 0 and 1, i.e. the random intercept explains either almost none or almost all of the variance, which is typically unrealistic.

The posterior (from the Gibbs sampler) is concentrated around \(p \approx 0.7\), indicating a substantial but not dominant contribution. It is much more concentrated and far from the extremes favored by the prior, suggesting the prior is not very reasonable.

```{r, echo=FALSE}
p_post <- samples[,23] / (samples[,23] + samples[,22])

curve(dbeta(x, 0.0001, 0.0001), from = 0, to = 1,
      main = "Prior and posterior of p",
      xlab = "p", ylab = "Density", lwd = 2)

lines(density(p_post), lwd = 2, lty = 2)
legend("top", legend = c("Prior", "Posterior"),
       lwd = 2, lty = c(1,2))
```
The posterior is concentrated around moderate values of \(p\), in contrast to the U-shaped prior. This shows that the data strongly update the prior and suggest a more realistic balance between the variance components.

## f) Analysis under the limiting prior \( \alpha \to 0 \)

To investigate the behaviour under the limiting prior, we rerun the Gibbs sampler with \(\alpha = 0\). The following trace plots illustrate the resulting Markov chains for the key parameters.

```{r, echo=FALSE}
samples_f <- gibbs_sampler(data, n_iter = 20000, alpha = 0)

par(mfrow = c(2,2))
plot(samples_f[,1], type = "l", main = expression(mu), ylab = expression(mu), xlab = "Iteration")
plot(samples_f[,22], type = "l", main = expression(1/tau), ylab = expression(1/tau), xlab = "Iteration")
plot(samples_f[,23], type = "l", main = expression(1/nu), ylab = expression(1/nu), xlab = "Iteration")
plot(log(samples_f[,23]), type = "l", main = expression(log(1/nu)),
     ylab = expression(log(1/nu)), xlab = "Iteration")
```

In the limiting case $\alpha \to 0$, the Gamma priors in (4) become the improper scale priors
\[
\pi(\tau)\propto \frac{1}{\tau}, \qquad \pi(\nu)\propto \frac{1}{\nu}.
\]

As in part (d), the chains for $\mu$ and $1/\tau$ remain stable and well-behaved, with similar posterior behavior.

In contrast, the random intercept variance $1/\nu$ shows substantially larger variability, with frequent large spikes. Its posterior is strongly right-skewed, indicating that very large values occur with non-negligible probability.

On the log scale, this appears as pronounced excursions corresponding to extreme values, highlighting that the posterior for $\nu$ is weakly constrained under the improper prior $\pi(\nu)\propto 1/\nu$.

Although the full conditional distributions remain proper, this behavior suggests that the joint posterior may be close to improper or poorly behaved. The variance component is therefore much less stable under this prior than under the Gamma prior used in part (d).


## g) Alternative prior based on total variance parametrization
Define
\[
\sigma^2=\nu^{-1}+\tau^{-1},
\qquad
p=\frac{\nu^{-1}}{\nu^{-1}+\tau^{-1}}.
\]
Then
\[
\nu^{-1}=p\sigma^2,\qquad \tau^{-1}=(1-p)\sigma^2,
\]
so that
\[
\nu=\frac{1}{p\sigma^2},\qquad \tau=\frac{1}{(1-p)\sigma^2}.
\]

To derive the implied prior on \((\nu,\tau)\), it is convenient to write
\[
\sigma^2=\frac{1}{\nu}+\frac{1}{\tau}=\frac{\nu+\tau}{\nu\tau},
\qquad
p=\frac{\nu^{-1}}{\nu^{-1}+\tau^{-1}}=\frac{\tau}{\nu+\tau}.
\]
The Jacobian of the transformation \((\nu,\tau)\mapsto (\sigma^2,p)\) is
\[
\frac{\partial(\sigma^2,p)}{\partial(\nu,\tau)}
=
\begin{pmatrix}
-\nu^{-2} & -\tau^{-2}\\[4pt]
-\dfrac{\tau}{(\nu+\tau)^2} & \dfrac{\nu}{(\nu+\tau)^2}
\end{pmatrix},
\]
with determinant
\[
\left|\frac{\partial(\sigma^2,p)}{\partial(\nu,\tau)}\right|
=
\frac{1}{\nu\tau(\nu+\tau)}.
\]

Since the prior in (7a) is
\[
\pi(\sigma^2,p)\propto \frac{1}{\sigma^2}I(0<p<1)I(\sigma^2>0),
\]
and
\[
\frac{1}{\sigma^2}=\frac{\nu\tau}{\nu+\tau},
\]
it follows by change of variables that
\[
\pi(\nu,\tau)
\propto
\frac{\nu\tau}{\nu+\tau}\cdot \frac{1}{\nu\tau(\nu+\tau)}
=
\frac{1}{(\nu+\tau)^2},
\]
for \(\nu>0\) and \(\tau>0\). Hence
\[
\pi(\nu,\tau)\propto \frac{1}{(\nu+\tau)^2}I(\nu>0)I(\tau>0).
\]

Next, fixing \(\nu\), the conditional prior density of \(\tau\) is
\[
\pi(\tau\mid \nu)\propto \frac{1}{(\nu+\tau)^2}, \qquad \tau>0.
\]
Let \(x=\ln\tau\), so that \(\tau=e^x\) and \(d\tau/dx=e^x\). Then
\[
\pi(x\mid \nu)\propto \frac{e^x}{(\nu+e^x)^2}
= \frac{e^{x-\ln\nu}}{(1+e^{x-\ln\nu})^2}.
\]
This is the density of a logistic distribution with location parameter \(\ln\nu\) and scale 1. Therefore,
\[
\ln\tau\mid \ln\nu \sim \text{Logistic}(\ln\nu,1).
\]

## h) Modification of the Gibbs sampler for the alternative prior
To sample from the posterior under the alternative prior in (7), we modify the Gibbs sampler from part (c). While the updates for \(\mu\) and \(\gamma_i\) remain unchanged, the parameters \(\nu\) and \(\tau\) are now updated using Metropolis–Hastings steps within a Gibbs framework.
```{r, echo=FALSE}
gibbs_sampler_prior7 <- function(data, n_iter = 20000,
                                 mu_init = NULL, gamma_init = NULL,
                                 tau_init = 1, nu_init = 1) {

  y <- data$y
  i <- as.numeric(factor(data$i))

  m <- length(unique(i))
  N <- length(y)

  mu <- if (is.null(mu_init)) mean(y) else mu_init
  gamma <- if (is.null(gamma_init)) rep(0, m) else gamma_init
  tau <- tau_init
  nu <- nu_init

  samples <- matrix(NA, n_iter, m + 3)
  colnames(samples) <- c("mu", paste0("gamma_", 1:m), "sigma2_eps", "sigma2_gamma")

  accept_tau <- 0
  accept_nu <- 0

  for (iter in 1:n_iter) {
    y_gamma <- y - gamma[i]
    mu <- rnorm(1, mean(y_gamma), sqrt(1 / (N * tau)))

    for (k in 1:m) {
      y_k <- y[i == k]
      n_k <- length(y_k)

      var_gamma <- 1 / (nu + n_k * tau)
      mean_gamma <- (tau * sum(y_k - mu)) / (nu + n_k * tau)

      gamma[k] <- rnorm(1, mean_gamma, sqrt(var_gamma))
    }

    sse <- sum((y - mu - gamma[i])^2)

    tau_prop <- rgamma(1, shape = N / 2, rate = sse / 2)

    log_acc_tau <- log(tau_prop) - 2 * log(nu + tau_prop) -
                   log(tau) + 2 * log(nu + tau)

    if (log(runif(1)) < log_acc_tau) {
      tau <- tau_prop
      accept_tau <- accept_tau + 1
    }
    
    ss_gamma <- sum(gamma^2)
    nu_prop <- rgamma(1, shape = m / 2, rate = ss_gamma / 2)

    log_acc_nu <- log(nu_prop) - 2 * log(nu_prop + tau) -
                  log(nu) + 2 * log(nu + tau)

    if (log(runif(1)) < log_acc_nu) {
      nu <- nu_prop
      accept_nu <- accept_nu + 1
    }
    samples[iter, ] <- c(mu, gamma, 1 / tau, 1 / nu)
  }

  list(
    samples = samples,
    accept_tau = accept_tau / n_iter,
    accept_nu = accept_nu / n_iter
  )
}
```


```{r,echo=FALSE}
out_h <- gibbs_sampler_prior7(data, 20000)
samples_h <- out_h$samples
```

```{r}
out_h$accept_tau
out_h$accept_nu
```

```{r, echo=FALSE}
par(mfrow = c(2,2))
plot(samples_h[,1], type = "l", main = expression(mu), ylab = expression(mu), xlab = "Iteration")
plot(samples_h[,22], type = "l", main = expression(1/tau), ylab = expression(1/tau), xlab = "Iteration")
plot(samples_h[,23], type = "l", main = expression(1/nu), ylab = expression(1/nu), xlab = "Iteration")
plot(log(samples_h[,23]), type = "l", main = expression(log(1/nu)),
     ylab = expression(log(1/nu)), xlab = "Iteration")
```

```{r, echo=FALSE}
p_h <- samples_h[,23] / (samples_h[,23] + samples_h[,22])
plot(density(p_h), main = "Posterior of p under prior (7)", xlab = "p")
```
To sample from the posterior under the prior in (7), we implemented a Metropolis-within-Gibbs sampler. The updates for $\mu$ and $\gamma_i$ remain unchanged from part (c), since their full conditional distributions are still normal. However, under the prior
\[
\pi(\nu,\tau)\propto \frac{1}{(\nu+\tau)^2},
\]
the full conditional distributions for $\tau$ and $\nu$ are no longer Gamma distributions. Therefore, these parameters were updated using Metropolis–Hastings steps.

As proposal distributions, we used the Gamma full conditionals derived in part (f):
\[
\tau' \sim \Gamma\left(\frac{N}{2},\frac12\sum (y-\mu-\gamma_i)^2\right),
\qquad
\nu' \sim \Gamma\left(\frac{m}{2},\frac12\sum \gamma_i^2\right).
\]
This results in acceptance probabilities
\[
A_\tau =
\min\!\left(
1,
\frac{\tau'}{(\nu+\tau')^2}\frac{(\nu+\tau)^2}{\tau}
\right),
\]
and
\[
A_\nu =
\min\!\left(
1,
\frac{\nu'}{(\nu'+\tau)^2}\frac{(\nu+\tau)^2}{\nu}
\right).
\]

Running the sampler for 20000 iterations produced acceptance rates of approximately $0.97$ for $\tau$ and $0.92$ for $\nu$, indicating that the proposal distributions lead to very efficient updates.

The trace plots show that the chains for $\mu$ and $1/\tau$ remain stable and well-behaved, similar to the earlier results. Importantly, the behavior of the random intercept variance $1/\nu$ is now considerably more stable than in part (f). The trace plot shows smaller fluctuations and no extreme excursions, and the trace of $\log(1/\nu)$ appears well-mixed without large spikes. This indicates improved convergence properties under the new prior.

The posterior distribution of the variance proportion
\[
p = \frac{\nu^{-1}}{\nu^{-1} + \tau^{-1}}
\]
is concentrated around moderate values, suggesting that a substantial but not dominant fraction of the total variance is explained by the random intercept. Unlike the earlier prior, the prior in (7) induces a uniform prior on $p$, avoiding the strong concentration near 0 and 1 that occurred under the Gamma prior in part (e). Consequently, the posterior under this prior appears more stable and interpretable.

Overall, the Gibbs sampler with the prior in (7) shows better convergence properties than the sampler using the improper scale prior in part (f), particularly for the random intercept variance component.

# Problem 3
## a) Stationarity properties of the AR(1) model

Consider the AR(1) model
$$
y_t - \mu = \phi (y_{t-1} - \mu) + \varepsilon_t, \qquad \varepsilon_t \sim N(0,\tau^{-1}),
$$
with $|\phi|<1$.

**Mean:**  
Taking expectations,
$$
E(y_t-\mu)=\phi E(y_{t-1}-\mu).
$$
Under stationarity, $E(y_t-\mu)=m$ is constant, so
$$
m=\phi m \;\Rightarrow\; (1-\phi)m=0.
$$
Since $|\phi|<1$, we get $m=0$, hence
$$
E(y_t \mid \mu,\phi,\tau)=\mu.
$$

**Variance:**  
Let $x_t=y_t-\mu$. Then
$$
x_t=\phi x_{t-1}+\varepsilon_t.
$$
Taking variances,
$$
\operatorname{Var}(x_t)=\phi^2 \operatorname{Var}(x_{t-1})+\tau^{-1}.
$$
Under stationarity, $\operatorname{Var}(x_t)=v$, so
$$
v=\phi^2 v+\tau^{-1}
\;\Rightarrow\;
v=\frac{\tau^{-1}}{1-\phi^2}.
$$

Thus,
$$
E(y_1 \mid \mu,\phi,\tau)=\mu, \qquad
\operatorname{Var}(y_1 \mid \mu,\phi,\tau)=\frac{\tau^{-1}}{1-\phi^2}.
$$


## b) Reparameterization and posterior derivation

Rewrite the model as
$$
y_t = \beta_0 + \beta_1 y_{t-1} + \varepsilon_t,
\qquad \varepsilon_t \sim N(0,\tau^{-1}),
$$
where
$$
\beta_1=\phi, \qquad \beta_0=(1-\phi)\mu.
$$
Hence,
$$
\mu=\frac{\beta_0}{1-\beta_1}, \qquad \phi=\beta_1.
$$

The prior on $(\mu,\phi,\tau)$ is
$$
\pi(\mu,\phi,\tau)\propto I(-1<\phi<1)I(\tau>0)\frac{1}{\tau}.
$$
Under the reparametrization, this becomes
$$
\pi(\beta_0,\beta_1,\tau)
\propto I(-1<\beta_1<1)I(\tau>0)\left|\frac{\partial(\mu,\phi)}{\partial(\beta_0,\beta_1)}\right|\frac{1}{\tau}.
$$

Since
$$
\mu=\frac{\beta_0}{1-\beta_1}, \qquad \phi=\beta_1,
$$
the Jacobian is
$$
\frac{\partial(\mu,\phi)}{\partial(\beta_0,\beta_1)}
=
\begin{pmatrix}
\frac{1}{1-\beta_1} & \frac{\beta_0}{(1-\beta_1)^2} \\
0 & 1
\end{pmatrix},
$$
so
$$
\left|\frac{\partial(\mu,\phi)}{\partial(\beta_0,\beta_1)}\right|
=\frac{1}{1-\beta_1}.
$$

Therefore,
$$
\pi(\beta_0,\beta_1,\tau)
\propto I(-1<\beta_1<1)I(\tau>0)\frac{1}{1-\beta_1}\frac{1}{\tau}.
$$

Now let
$$
\beta=
\begin{pmatrix}
\beta_0\\
\beta_1
\end{pmatrix},
\qquad
X=
\begin{pmatrix}
1 & y_1\\
1 & y_2\\
\vdots & \vdots\\
1 & y_{n-1}
\end{pmatrix},
$$
so that
$$
y_{2:n}\mid \beta,\tau \sim N(X\beta,\tau^{-1}I).
$$

Thus the likelihood contribution from $y_2,\dots,y_n$ is
$$
\tau^{(n-1)/2}\exp\left(-\frac{\tau}{2}(y_{2:n}-X\beta)^T(y_{2:n}-X\beta)\right).
$$

From part (a), stationarity implies
$$
y_1\mid \beta,\tau \sim N\left(\frac{\beta_0}{1-\beta_1},\frac{\tau^{-1}}{1-\beta_1^2}\right),
$$
so its density is proportional to
$$
\sqrt{1-\beta_1^2}\,
\exp\left(
-\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\right).
$$

Combining likelihood and prior gives
$$
\pi(\beta,\tau\mid y)
\propto
\tau^{n/2-1}
\exp\left(-\frac{\tau}{2}(y_{2:n}-X\beta)^T(y_{2:n}-X\beta)\right)
\frac{\sqrt{1-\beta_1^2}\,I(-1<\beta_1<1)}{1-\beta_1}
\exp\left(
-\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\right).
$$

## c) Metropolis–Hastings acceptance probability

Let
$$
Q(\beta',\tau' \mid \beta,\tau,y)\propto
\tau'^{\,n/2-1}
\exp\left(-\frac{\tau'}{2}(y_{2:n}-X\beta')^T(y_{2:n}-X\beta')\right)
$$
denote the independence proposal based on the part marked $(*)$ in (b).

From part (b), the target posterior is
$$
\pi(\beta,\tau\mid y)\propto
Q(\beta,\tau\mid \beta',\tau',y)\,
\frac{\sqrt{1-\beta_1^2}\,I(-1<\beta_1<1)}{1-\beta_1}
\exp\left(
-\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\right).
$$

Since the proposal is an independence proposal, the Metropolis--Hastings ratio is
$$
r(\beta',\tau' \mid \beta,\tau)
=
\frac{\pi(\beta',\tau'\mid y)\,Q(\beta,\tau\mid \beta',\tau',y)}
{\pi(\beta,\tau\mid y)\,Q(\beta',\tau'\mid \beta,\tau,y)}.
$$

The $Q$-terms cancel, so
$$
r(\beta',\tau' \mid \beta,\tau)
=
\frac{\frac{\sqrt{1-(\beta_1')^2}\,I(-1<\beta_1'<1)}{1-\beta_1'}
\exp\left(
-\frac{\tau'}{2}(1-(\beta_1')^2)\left(y_1-\frac{\beta_0'}{1-\beta_1'}\right)^2
\right)}
{\frac{\sqrt{1-\beta_1^2}\,I(-1<\beta_1<1)}{1-\beta_1}
\exp\left(
-\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\right)}.
$$

Hence the log acceptance probability is
$$
\log \alpha(\beta',\tau' \mid \beta,\tau,y)
=
\min\Bigg\{0,\,
\frac{1}{2}\log\big(1-(\beta_1')^2\big)-\log(1-\beta_1')
-\frac{\tau'}{2}(1-(\beta_1')^2)\left(y_1-\frac{\beta_0'}{1-\beta_1'}\right)^2
$$
$$
\qquad\qquad
-\frac{1}{2}\log(1-\beta_1^2)+\log(1-\beta_1)
+\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\Bigg\},
$$
for $-1<\beta_1,\beta_1'<1$.

Equivalently,
$$
\log \alpha(\beta',\tau' \mid \beta,\tau,y)
=
\min\left\{0,\,
\log\frac{\sqrt{1-(\beta_1')^2}/(1-\beta_1')}
{\sqrt{1-\beta_1^2}/(1-\beta_1)}
-\frac{\tau'}{2}(1-(\beta_1')^2)\left(y_1-\frac{\beta_0'}{1-\beta_1'}\right)^2
+\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2
\right\}.
$$

## d) Factorization and proposal distribution

Recall that
$$
(*) \;=\;
\tau^{n/2-1}\exp\left(-\frac{\tau}{2}(y_{2:n}-X\beta)^T(y_{2:n}-X\beta)\right).
$$

Let
$$
Q=X^TX,\qquad
\hat\beta=(X^TX)^{-1}X^Ty_{2:n},\qquad
\hat\varepsilon=y_{2:n}-X\hat\beta.
$$

Using the usual least-squares decomposition,
$$
(y_{2:n}-X\beta)^T(y_{2:n}-X\beta)
=
(\beta-\hat\beta)^TQ(\beta-\hat\beta)+\hat\varepsilon^T\hat\varepsilon.
$$

Hence
$$
(*) \propto
\tau^{n/2-1}
\exp\left(-\frac{\tau}{2}(\beta-\hat\beta)^TQ(\beta-\hat\beta)\right)
\exp\left(-\frac{\tau}{2}\hat\varepsilon^T\hat\varepsilon\right).
$$

Since
$$
\tau^{n/2-1}=\tau^{p/2}\tau^{(n-p)/2-1},
$$
we can write
$$
(*) \propto
|Q|^{1/2}\tau^{p/2}
\exp\left(-\frac{\tau}{2}(\beta-\hat\beta)^TQ(\beta-\hat\beta)\right)
\tau^{(n-p)/2-1}
\exp\left(-\frac{\tau}{2}\hat\varepsilon^T\hat\varepsilon\right).
$$

Therefore, up to proportionality,
$$
\beta \mid \tau,y \sim N\!\left(\hat\beta,\;(\tau Q)^{-1}\right),
$$
and
$$
\tau \mid y \sim \Gamma\!\left(\frac{n-p}{2},\;\frac{1}{2}\hat\varepsilon^T\hat\varepsilon\right),
$$
where we use the shape--rate parameterization.

Thus, to simulate from the joint proposal in part (c), we can first draw
$$
\tau \sim \Gamma\!\left(\frac{n-p}{2},\;\frac{1}{2}\hat\varepsilon^T\hat\varepsilon\right),
$$
and then draw
$$
\beta \mid \tau \sim N\!\left(\hat\beta,\;(\tau X^TX)^{-1}\right).
$$

## e) Implementation of the Metropolis–Hastings sampler

We use an independence Metropolis--Hastings sampler with proposal from part (d):
$$
\tau' \sim \Gamma\left(\frac{n-p}{2}, \frac{1}{2}\hat\varepsilon^T\hat\varepsilon\right),
\qquad
\beta' \mid \tau' \sim N\left(\hat\beta, (\tau' X^TX)^{-1}\right),
$$
with $p=2$.

The log acceptance ratio from part (c) is
$$
\log r
=
\log\frac{\sqrt{1-(\beta_1')^2}/(1-\beta_1')}
{\sqrt{1-\beta_1^2}/(1-\beta_1)}
-\frac{\tau'}{2}(1-(\beta_1')^2)\left(y_1-\frac{\beta_0'}{1-\beta_1'}\right)^2
+\frac{\tau}{2}(1-\beta_1^2)\left(y_1-\frac{\beta_0}{1-\beta_1}\right)^2,
$$
provided $-1<\beta_1,\beta_1'<1$. Otherwise the proposal is rejected.

```{r, echo=FALSE}
mh_ar1 <- function(y, n_iter = 20000, beta_init = NULL, tau_init = 1) {
  n <- length(y)
  y_resp <- y[2:n]
  X <- cbind(1, y[1:(n-1)])
  p <- ncol(X)
  
  Q <- crossprod(X)
  Q_inv <- solve(Q)
  beta_hat <- Q_inv %*% crossprod(X, y_resp)
  beta_hat <- as.numeric(beta_hat)
  eps_hat <- y_resp - X %*% beta_hat
  sse_hat <- sum(eps_hat^2)
  
  shape_prop <- (n - p) / 2
  rate_prop  <- sse_hat / 2
  
  if (is.null(beta_init)) beta_init <- beta_hat
  
  beta_chain <- matrix(NA_real_, n_iter, 2)
  tau_chain  <- numeric(n_iter)
  accept     <- logical(n_iter)
  
  beta_chain[1, ] <- beta_init
  tau_chain[1] <- tau_init
  
  log_target_extra <- function(beta, tau, y1) {
    b0 <- beta[1]
    b1 <- beta[2]
    
    if (b1 <= -1 || b1 >= 1 || tau <= 0) return(-Inf)
    
    log(sqrt(1 - b1^2) / (1 - b1)) -
      0.5 * tau * (1 - b1^2) * (y1 - b0 / (1 - b1))^2
  }
  
  for (iter in 2:n_iter) {
    tau_prop <- rgamma(1, shape = shape_prop, rate = rate_prop)
    beta_prop <- as.numeric(rmvnorm(1, mean = beta_hat, sigma = Q_inv / tau_prop))
    
    log_r <- log_target_extra(beta_prop, tau_prop, y[1]) -
             log_target_extra(beta_chain[iter - 1, ], tau_chain[iter - 1], y[1])
    
    if (log(runif(1)) < log_r) {
      beta_chain[iter, ] <- beta_prop
      tau_chain[iter] <- tau_prop
      accept[iter] <- TRUE
    } else {
      beta_chain[iter, ] <- beta_chain[iter - 1, ]
      tau_chain[iter] <- tau_chain[iter - 1]
    }
  }
  
  out <- cbind(beta0 = beta_chain[,1],
               beta1 = beta_chain[,2],
               tau = tau_chain)
  
  attr(out, "accept_rate") <- mean(accept[-1])
  mcmc(out)
}
```

```{r, echo=FALSE}
load(url("https://www.math.ntnu.no/emner/TMA4300/2026v/clouds.RData"))

set.seed(1)
mcmc_ar1 <- mh_ar1(clouds$z, n_iter = 20000)

attr(mcmc_ar1, "accept_rate")
summary(mcmc_ar1)
plot(mcmc_ar1)
```
The trace plots for $\beta_0$, $\beta_1$, and $\tau$ show stable behavior with no visible trends, indicating good mixing and convergence of the chains. The samples fluctuate around constant levels, and the corresponding posterior densities appear smooth and unimodal. The high acceptance rate ($\approx 0.92$) further suggests that the proposal distribution is well chosen and leads to efficient sampling.

## f) Conditional distribution under censoring

For an interior time point $j$, the AR(1) model implies
$$
y_j \mid y_{-j}, \beta, \tau
\sim
N\left(
\mu + \frac{\phi}{1+\phi^2}(y_{j-1}+y_{j+1}-2\mu),
\frac{\tau^{-1}}{1+\phi^2}
\right),
$$
where
$$
\mu=\frac{\beta_0}{1-\beta_1}, \qquad \phi=\beta_1.
$$

Now condition also on the observed data $(z,\delta)$.

- If $\delta_j=1$, then $y_j=z_j$ is observed exactly, so
$$
y_j \mid y_{-j},\delta,z,\beta,\tau
=
z_j
\quad \text{with probability 1.}
$$

- If $\delta_j=0$, then the observation is right-censored, so $y_j > z_j$. Hence
$$
y_j \mid y_{-j},\delta,z,\beta,\tau
\sim
N\left(m_j,\;s_j^2\right)\,I(y_j>z_j),
$$
that is, a lower-truncated normal distribution, where
$$
m_j
=
\mu + \frac{\phi}{1+\phi^2}(y_{j-1}+y_{j+1}-2\mu),
\qquad
s_j^2
=
\frac{\tau^{-1}}{1+\phi^2}.
$$

Equivalently, its density is
$$
\pi(y_j \mid y_{-j},\delta,z,\beta,\tau)
=
\frac{
\frac{1}{s_j}\varphi\!\left(\frac{y_j-m_j}{s_j}\right)
}{
1-\Phi\!\left(\frac{z_j-m_j}{s_j}\right)
}
\,I(y_j>z_j),
$$
for $\delta_j=0$, where $\varphi$ and $\Phi$ are the standard normal density and distribution function.

## g) Gibbs sampling with censored data

We now account for right-censoring by alternating between:

1. updating $(\beta,\tau)$ using the independence Metropolis--Hastings step from part (e), conditional on the current completed series $y$;
2. updating each censored $y_j$ from its full conditional distribution in part (f).

Thus this is a Gibbs sampler with an MH step for $(\beta,\tau)$.

```{r, echo=FALSE}
rtruncnorm_lower <- function(n, mean = 0, sd = 1, lower = -Inf) {
  u <- runif(n, pnorm(lower, mean = mean, sd = sd), 1)
  qnorm(u, mean = mean, sd = sd)
}

clouds_gibbs <- function(z, delta, n_iter = 20000, beta_init = c(0, 0.9), tau_init = 1) {
  n <- length(z)
    y <- z
  cens_idx <- which(delta == 0)
  y[cens_idx] <- z[cens_idx] + 0.1
  
  beta <- beta_init
  tau <- tau_init
  
  out <- matrix(NA_real_, n_iter, 3 + length(cens_idx))
  colnames(out) <- c("beta0", "beta1", "tau", paste0("y_cens_", cens_idx))
  
  accept <- logical(n_iter)
  
  log_target_extra <- function(beta, tau, y1) {
    b0 <- beta[1]
    b1 <- beta[2]
    if (b1 <= -1 || b1 >= 1 || tau <= 0) return(-Inf)
    
    log(sqrt(1 - b1^2) / (1 - b1)) -
      0.5 * tau * (1 - b1^2) * (y1 - b0 / (1 - b1))^2
  }
  
  for (iter in 1:n_iter) {
    y_resp <- y[2:n]
    X <- cbind(1, y[1:(n-1)])
    p <- ncol(X)
    
    Q <- crossprod(X)
    Q_inv <- solve(Q)
    beta_hat <- as.numeric(Q_inv %*% crossprod(X, y_resp))
    eps_hat <- y_resp - X %*% beta_hat
    sse_hat <- sum(eps_hat^2)
    
    shape_prop <- (n - p) / 2
    rate_prop  <- sse_hat / 2
    
    tau_prop <- rgamma(1, shape = shape_prop, rate = rate_prop)
    beta_prop <- beta_hat + t(chol(Q_inv / tau_prop)) %*% rnorm(2)
    beta_prop <- as.numeric(beta_prop)
    
    log_r <- log_target_extra(beta_prop, tau_prop, y[1]) -
             log_target_extra(beta, tau, y[1])
    
    if (log(runif(1)) < log_r) {
      beta <- beta_prop
      tau <- tau_prop
      accept[iter] <- TRUE
    }
    
    b0 <- beta[1]
    b1 <- beta[2]
    mu <- b0 / (1 - b1)
    phi <- b1
    
    for (j in cens_idx) {
      if (j == 1) {
        mean_j <- mu + phi * (y[2] - mu)
        sd_j <- sqrt(1 / tau)
      } else if (j == n) {
        mean_j <- mu + phi * (y[n - 1] - mu)
        sd_j <- sqrt(1 / tau)
      } else {
        mean_j <- mu + (phi / (1 + phi^2)) * (y[j - 1] + y[j + 1] - 2 * mu)
        sd_j <- sqrt((1 / tau) / (1 + phi^2))
      }
      
      y[j] <- rtruncnorm_lower(1, mean = mean_j, sd = sd_j, lower = z[j])
    }
    
    out[iter, ] <- c(beta, tau, y[cens_idx])
  }
  
  attr(out, "accept_rate") <- mean(accept)
  attr(out, "censored_index") <- cens_idx
  mcmc(out)
}
```


```{r, echo=FALSE}
load(url("https://www.math.ntnu.no/emner/TMA4300/2026v/clouds.RData"))

set.seed(1)
mcmc_clouds <- clouds_gibbs(clouds$z, clouds$delta, n_iter = 20000)

attr(mcmc_clouds, "accept_rate")
summary(mcmc_clouds[, 1:3])
effectiveSize(mcmc_clouds[, 1:3])

plot(mcmc_clouds[, 1:3])
```

```{r, echo=FALSE}
cens_idx <- attr(mcmc_clouds, "censored_index")
draw_idx <- seq(5000, 20000, length.out = 5)

plot(clouds$z, type = "l", col = "grey", xlab = "Time", ylab = "log(cloud ceiling height)")
for (k in draw_idx) {
  y_draw <- clouds$z
  y_draw[cens_idx] <- as.numeric(mcmc_clouds[k, -(1:3)])
  lines(y_draw)
}
```
The trace plots for $\beta_0$, $\beta_1$, and $\tau$ indicate stable behavior with no visible trends, suggesting that the Gibbs sampler converges satisfactorily even when accounting for censoring. The chains fluctuate around constant levels, and the posterior densities remain smooth.

The sampled trajectories of $y$ exhibit substantial variability over time, reflecting the uncertainty introduced by the censored observations. Occasional large drops correspond to imputed values for censored data points, showing that the model allows for a wide range of plausible realizations.

Overall, the sampler appears to perform well under the extended model, although convergence is somewhat slower compared to the uncensored case due to the additional uncertainty from the latent variables.

# Problem 4

We consider the state equation
\begin{equation}
u_t = u_{t-1} + r \left(1 - \left(\frac{e^{u_{t-1}}}{K}\right)^\theta \right) + e_t,
\qquad e_t \sim \mathcal{N}(0,\sigma),
\end{equation}
and the observation equation
\begin{equation}
y_t \mid u_t \sim \text{Poisson}(c e^{u_t}).
\end{equation}

## a) Non-identifiability and identifiability constraint

To show that the model is non-identifiable, it suffices to demonstrate that different parameter values yield the same distribution for the observations $y = (y_1,\dots,y_n)$.

Let $a \in \mathbb{R}$ and define
\[
u_t^* = u_t + a, 
\qquad K^* = K e^a, 
\qquad c^* = c e^{-a},
\]
with
\[
r^* = r, \qquad \theta^* = \theta, \qquad \sigma^* = \sigma.
\]

Then
\[
\frac{e^{u_{t-1}^*}}{K^*}
= \frac{e^{u_{t-1}+a}}{K e^a}
= \frac{e^{u_{t-1}}}{K},
\]
so the state equation remains unchanged:
\[
u_t^* = u_{t-1}^* + r \left(1 - \left(\frac{e^{u_{t-1}^*}}{K^*}\right)^\theta \right) + e_t.
\]

For the observation model,
\[
c^* e^{u_t^*} = (c e^{-a}) e^{u_t + a} = c e^{u_t},
\]
so the conditional distribution of \(y_t\) is unchanged.

Since \(u_1\) has an improper uniform distribution on \(\mathbb{R}\), it is invariant under shifts. Hence, the transformation
\[
(u_t, K, c) \mapsto (u_t + a, K e^a, c e^{-a})
\]
leaves the likelihood unchanged, and the model is non-identifiable.

\paragraph{Identifiability.}
The non-identifiability arises from a shift invariance in \(u_t\). The quantity
\[
\log K + \log c
\]
is invariant under the transformation, implying that only the product \(cK\) is identifiable. To remove this invariance, we impose the constraint
\[
c = 1,
\]
which yields an identifiable model without altering its structure.

## b)Joint negative log-likelihood and RTMB implementation

To resolve the non-identifiability shown in part (a), we fix \(c = 1\). The parameters \(r, K, \theta, \sigma\) are estimated, with \(K\), \(\theta\), and \(\sigma\) optimized on the log-scale to enforce positivity.

We write the negative joint log-likelihood as
\[
-\log f(y,u \mid \theta)
= -\sum_{t=1}^n \log p(y_t \mid u_t)
  - \sum_{t=2}^n \log p(u_t \mid u_{t-1}),
\]
where

\paragraph{Observation model}
\[
y_t \mid u_t \sim \text{Poisson}(e^{u_t}),
\]
so
\[
-\log p(y_t \mid u_t)
= e^{u_t} - y_t u_t + \log(y_t!).
\]

\paragraph{State model}
\[
u_t \mid u_{t-1} \sim \mathcal{N}(\mu_t, \sigma^2),
\quad
\mu_t = u_{t-1} + r\left(1 - \left(\frac{e^{u_{t-1}}}{K}\right)^\theta\right),
\]
giving
\[
-\log p(u_t \mid u_{t-1})
=
\frac{(u_t - \mu_t)^2}{2\sigma^2} + \log \sigma + \text{const}.
\]

The latent states \(u = (u_1,\dots,u_n)\) are treated as random effects in \texttt{RTMB}. The marginal likelihood is approximated using the Laplace approximation via \texttt{MakeADFun} with \texttt{random = "u"}. Parameters \(K\), \(\theta\), and \(\sigma\) are optimized on the log-scale to enforce positivity, and \(c=1\) is fixed to ensure identifiability.

```{r model-fit, echo=TRUE}
load(url("https://www.math.ntnu.no/emner/TMA4300/2026v/theta-logistic.RData"))

f <- function(parms) {
  r     <- exp(parms$log_r)
  K     <- exp(parms$log_K)
  theta <- exp(parms$log_theta)
  sigma <- exp(parms$log_sigma)
  
  u <- parms$u
  n <- length(y)
  
  nll <- 0
  for (t in 2:n) {
    mu_u <- u[t - 1] + r * (1 - (exp(u[t - 1]) / K)^theta)
    nll <- nll - dnorm(u[t], mean = mu_u, sd = sigma, log = TRUE)
  }
  
  for (t in 1:n) {
    lambda <- exp(u[t])
    nll <- nll - dpois(y[t], lambda = lambda, log = TRUE)
  }
    ADREPORT(r)
  ADREPORT(K)
  ADREPORT(theta)
  ADREPORT(sigma)
  
  nll
}

parameters <- list(
  log_r = 0,
  log_K = log(mean(y)),
  log_theta = 0,
  log_sigma = -1,
  u = log(y + 1)
)

obj <- MakeADFun(f, parameters, random = "u")
```

## c) Parameter estimation and latent state reconstruction

The Laplace approximation of the marginal likelihood was maximised using the \texttt{nlminb} optimizer. The optimisation converged successfully, with convergence code 0 and a message indicating both parameter and relative convergence. The maximum absolute gradient at the optimum was $8.96 \cdot 10^{-7}$, which is close to zero and confirms that a stationary point was reached.

```{r, echo=FALSE, results= 'hide', message=FALSE, warning=FALSE}
opt <- nlminb(obj$par, obj$fn, obj$gr)

sdr <- sdreport(obj)

u_ests <- summary(sdr, "random")[, 1]
u_ses  <- summary(sdr, "random")[, 2]

par_est <- summary(sdr, "fixed")

grad_max <- max(abs(obj$gr(opt$par)))
conv <- opt$convergence
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
opt$convergence      # 0 = success
opt$message          # should say something like "relative convergence"
opt$objective
max(abs(obj$gr(opt$par)))  # gradient size
```

Parameter estimates and standard errors were obtained using \texttt{sdreport}, which is based on a Gaussian approximation around the optimum using the curvature of the objective function. In particular, \texttt{sdreport} provides both the estimated latent states and their uncertainties.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.pos="H"}

par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

# Plot latent states
plot(u_ests, type = "l", col = "blue", lwd = 2,
     main = "Estimated latent log-population dynamics", ylab = "Log-population u_t", xlab = "Time")
lines(u_ests + 1.96 * u_ses, col = "blue", lty = 2)
lines(u_ests - 1.96 * u_ses, col = "blue", lty = 2)

# Plot observations vs fitted
plot(y, type = "p", pch = 16, col = "gray",
     main = "Observed counts y_t and model-based expected counts exp(u_t)",
     ylab = "Counts", xlab = "Time")
lines(exp(u_ests), col = "red", lwd = 2)

legend("topright",
       legend = c("y_t", "exp(u_t)"),
       col = c("gray", "red"),
       pch = c(16, NA), lty = c(NA, 1), bty = "n")

```


The estimated latent states $\hat{u}_t$ are shown in the first plot. Since $u_t$ represents the latent log-population, this plot describes the model’s estimate of the underlying population dynamics on the log-scale. The trajectory is smooth over time because the latent states are constrained by the theta-logistic state model, and the confidence bands are relatively narrow, suggesting that the latent process is well informed by the data.

The second plot compares the observed counts $y_t$ with the fitted values $\exp(\hat{u}_t)$, which represent the model-based expected counts on the original scale. The fitted curve follows the observed data closely, including periods of rapid increase and decline, showing that the model captures the main structure of the data well. At the same time, the fitted values are not obtained by fitting each time point independently, but through the latent state process, which imposes temporal dependence and yields a coherent trajectory over time.

Overall, the model provides a good fit and successfully recovers an underlying latent population process from noisy count observations.

## d) Simulation study and standard error validation

To evaluate the performance of the method, we simulated a dataset of \(n=200\) observations using known parameter values and fitted the model to the simulated data.

```{r, echo=FALSE, results = 'hide', message=FALSE, warning=FALSE}
simulate_theta_logistic <- function(n, r, K, theta, sigma, c = 1, u1 = 0) {
  u <- numeric(n)
  y <- numeric(n)
  u[1] <- u1
  
  for (t in 2:n) {
    eps <- rnorm(1, 0, sigma)
    u[t] <- u[t-1] + r * (1 - (exp(u[t-1])/K)^theta) + eps
  }
  
  y <- rpois(n, c * exp(u))
  
  list(u = u, y = y)
}

set.seed(123)
true_pars <- list(r = 0.1, K = 45, theta = 2.0, sigma = 0.3)

sim_data <- simulate_theta_logistic(
  n = 200,
  r = true_pars$r,
  K = true_pars$K,
  theta = true_pars$theta,
  sigma = true_pars$sigma
)

y <- sim_data$y

parameters_sim <- list(
  log_r = 0,
  log_K = log(mean(y)),
  log_theta = 0,
  log_sigma = -1,
  u = log(y + 1)
)

obj_sim <- MakeADFun(f, parameters_sim, random = "u")
opt_sim <- nlminb(obj_sim$par, obj_sim$fn, obj_sim$gr)
sdr_sim <- sdreport(obj_sim)

summary_sim <- summary(sdr_sim, "report")
print(summary_sim)
```

### Results

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
estimates <- summary_sim[, "Estimate"]
ses       <- summary_sim[, "Std. Error"]
param_names <- rownames(summary_sim)

true_vals <- c(r = 0.1, K = 45, theta = 2.0, sigma = 0.3)

lower <- estimates - 1.96 * ses
upper <- estimates + 1.96 * ses

ci_table <- data.frame(
  Parameter = c("$r$", "$K$", "$\\theta$", "$\\sigma$"),
  "True value" = true_vals[param_names],
  Estimate = round(estimates, 4),
  "Std. error" = round(ses, 4),
  "95 % CI" = paste0("[", round(lower, 3), ", ", round(upper, 3), "]")
)

colnames(ci_table) <- c(
  "Parameter",
  "True value",
  "Estimate",
  "Std. error",
  "95\\% CI"
)

kable(ci_table,
      format = "latex",
      booktabs = TRUE,
      escape = FALSE,
      caption = "Comparison of true parameter values and estimates based on simulated data.") %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

The parameter estimates are close to the true values, and all true parameters lie within the corresponding 95\% confidence intervals. This indicates that the estimation procedure performs well and that the Laplace approximation provides reliable inference for this model.

The parameter $\theta$ exhibits the largest uncertainty, reflected in a relatively wide confidence interval. This is expected, as $\theta$ governs the nonlinear density dependence and enters the model in a highly nonlinear manner, making it more difficult to estimate from finite data. In contrast, $\sigma$ is estimated with high precision, indicating that the noise level is well identified. The parameters $r$ and $K$ are also reasonably well estimated, suggesting that the overall growth dynamics are captured effectively.

The results are based on a single simulated dataset, and the estimation performance may vary across different realizations. In addition, the Laplace approximation relies on a Gaussian approximation of the latent posterior distribution, which may be less accurate in highly nonlinear settings or for small sample sizes. Nevertheless, the results suggest that the approximation performs well in this setting.

Fixing $c = 1$ was essential to ensure identifiability. Without this constraint, the model exhibits a shift invariance in the latent states, making it impossible to uniquely determine the parameters. Imposing this constraint leads to a well-defined estimation problem and stable numerical optimization.

\textbf{Standard errors}

The standard errors reported by \texttt{sdreport} are based on asymptotic likelihood theory. The covariance matrix is approximated by the inverse Hessian of the Laplace-approximated marginal negative log-likelihood,
\[
H(\hat{\theta})^{-1}.
\]
Since parameters are optimized on the log-scale, the Delta method is used to obtain standard errors on the original scale:
\[
\mathrm{Var}(g(\theta)) \approx \nabla g(\theta)^T H(\theta)^{-1} \nabla g(\theta).
\]

Overall, the standard errors appear to be of a reasonable magnitude, and the agreement between estimated and true parameter values provides empirical support for the reliability of the Laplace approximation in this model.





