---
title: "Project 2 – TMA4300"
author: "Karine Grande & Thea Boge"
date: "2026-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Library imports
```{r}
library(glmmTMB)
```

# Problem 1
We suppose that the joint density of $U$, $Z$ up to a normalizing constant is 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]

## a) 
If a joint density is known up to proportionality, the full conditional of one variable is obtained by 
\[
f_{X|Y} (x|y) \propto f^{*}_{U,Z}(u,z) \textit{as a function of x}
\]

### Conditional ditribution of $U|Z=z$

Fix $z$. The joint density becomes 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]
As a function of $u$, this is constant on the interval $(0,e^{-z^2/2}})$. 

The normalizing constant is
\[
\int_0^{e^{-z^2/2}} 1 \, du =e^{-z^2/2}.
\]

Therefore,
\[
f_{U \mid Z}(u \mid z) = \frac{1}{e^{-z^2/2}} = e^{z^2/2},\quad 0 < u < e^{-z^2/2}.
\]

Hence,
\[
U \mid Z = z \sim \mathrm{Unif} \big(0, e^{-z^2/2}\big).
\]. 

### Conditional ditribution of $Z|U=u$
Fix $u$. The support condition is $0 < u < e^{-z^2/2}$.

Solving for $z$:
\[
u < e^{-z^2/2} \quad \Longleftrightarrow \quad \log u < -\frac{z^2}{2}
\]

\[
-2 \log u > z^2
\]

\[
|z| < \sqrt{-2 \log u}
\]

Define
\[
a(u) = \sqrt{-2 \log u}
\]

Therefore, the support of the conditional distribution is
\[
z \in (-a(u), a(u)).
\]

Within this interval the joint density is constant and equal to 1. The normalizing constant is therefore the length of the interval:
\[
\int_{-a(u)}^{a(u)} 1 \, dz = 2a(u).
\]

Thus, the full conditional density of \( Z \mid U = u \) is

\[
f_{Z \mid U}(z \mid u) = \frac{1}{2a(u)},\quad |z| < a(u),
\]

where
\[
a(u) = \sqrt{-2 \log u}.
\]

Thus,
\[
Z \mid U = u \sim \mathrm{Unif}\big(-a(u), a(u)\big).
\]

### Gibbs Sampler

Given an initial valid point \( (U^{(0)}, Z^{(0)}) \) satisfying
\( 0 < U^{(0)} < e^{-(Z^{(0)})^2/2} \),
iterate for \( t = 1, \dots, n \):

\begin{enumerate}
    \item Sample
    \[
    U^{(t)} \sim 
    \mathrm{Unif}\left(0, e^{-(Z^{(t-1)})^2/2}\right).
    \]

    \item Define
    \[
    a^{(t)} = \sqrt{-2 \log U^{(t)}}.
    \]

    \item Sample
    \[
    Z^{(t)} \sim 
    \mathrm{Unif}\left(-a^{(t)},\, a^{(t)}\right).
    \]
\end{enumerate}

This iterative scheme defines a Markov chain whose stationary distribution is the joint density \( f_{U,Z}(u,z) \).The marginal stationary distribution of \( Z \) is \( \mathcal{N}(0,1) \).

```{r}
gibbs_normal <- function(n, u_init = 0.5, z_init = 0) {
  
  U <- numeric(n)
  Z <- numeric(n)
  
  U[1] <- u_init
  Z[1] <- z_init
  
  for (t in 2:n) {
    
    # Sample U | Z
    U[t] <- runif(1, 0, exp(-(Z[t-1]^2)/2))
    
    # Compute truncation bound
    a <- sqrt(-2 * log(U[t]))
    
    # Sample Z | U
    Z[t] <- runif(1, -a, a)
  }
  
  Z  # return only Z since that is the target
}
```

```{r}
# Run Gibbs sampler
set.seed(123)
Z <- gibbs_normal(10000)

# Burn-in removal
Z_post <- Z[1001:10000]

# Diagnostics
mean(Z_post)
var(Z_post)

hist(Z_post, probability = TRUE, breaks = 40)
curve(dnorm(x), col="red", add=TRUE)
```
The posterior sample mean is approximately
\[
\hat{\mu} \approx -0.019,
\]
and the empirical variance is
\[
\hat{\sigma}^2 \approx 0.97.
\]

Both values are close to the theoretical mean and variance,
\[
\mathbb{E}[Z] = 0,\qquad \mathrm{Var}(Z) = 1,
\]
of a standard normal distribution.

The histogram aligns well with the standard normal density curve, confirming that the Gibbs sampler correctly targets
\[
Z \sim \mathcal{N}(0,1).
\]

## b)
We determine the marginal distribution of \( Z \) by integrating out \( U \).

Recall that the joint density (up to normalization) is

\[
f^*_{U,Z}(u,z) =
\begin{cases}
1, & 0 < u < e^{-z^2/2}, \\
0, & \text{otherwise}.
\end{cases}
\]

The marginal density of \( Z \) is therefore

\[
f_Z(z)
=
\int_0^{e^{-z^2/2}} 1 \, du
=
e^{-z^2/2}.
\]

Thus,

\[
f_Z(z) \propto e^{-z^2/2}.
\]


Since
\[
\int_{-\infty}^{\infty} e^{-z^2/2} \, dz
=
\sqrt{2\pi},
\]
the normalized density is

\[
f_Z(z)
=
\frac{1}{\sqrt{2\pi}} e^{-z^2/2}.
\]

Therefore,
\[
Z \sim \mathcal{N}(0,1).
\]
Hence, discarding the \( U \)-values from the Gibbs sampler yields a sample from a standard normal distribution.

## c)
### Correlation between successive Gibbs samples $Z_{n-1}$ and $Z_n$}

From the Gibbs sampler we have
\[
Z_n \mid U_n \sim \mathrm{Unif}\!\big(-a(U_n), a(U_n)\big),
\qquad
a(u) = \sqrt{-2 \log u}.
\]

This conditional distribution is symmetric around zero. Therefore,
\[
\mathbb{E}[Z_n \mid U_n] = 0.
\]

By the law of total expectation,
\[
\mathbb{E}[Z_n]
=
\mathbb{E}\!\left[\mathbb{E}[Z_n \mid U_n]\right]
=
\mathbb{E}[0]
=
0.
\]

Since the Gibbs sampler has stationary distribution $Z_n \sim \mathcal{N}(0,1)$, it follows that
\[
\mathbb{E}[Z_{n-1}] = \mathbb{E}[Z_n] = 0.
\]

We now compute the covariance. By definition,
\[
\mathrm{Cov}(Z_{n-1}, Z_n)
=
\mathbb{E}[Z_{n-1}Z_n]
-
\mathbb{E}[Z_{n-1}]\mathbb{E}[Z_n].
\]

Because both expectations are zero, this reduces to
\[
\mathrm{Cov}(Z_{n-1}, Z_n)
=
\mathbb{E}[Z_{n-1}Z_n].
\]

Using the tower property (law of iterated expectation),
\[
\mathbb{E}[Z_{n-1}Z_n]
=
\mathbb{E}\!\left[
\mathbb{E}[Z_{n-1}Z_n \mid U_n]
\right].
\]

In the Gibbs sampler, $Z_n$ is generated from $U_n$ independently of $Z_{n-1}$. Hence, conditional on $U_n$, the variables $Z_{n-1}$ and $Z_n$ are independent. Therefore,
\[
\mathbb{E}[Z_{n-1}Z_n \mid U_n]
=
Z_{n-1} \, \mathbb{E}[Z_n \mid U_n].
\]

Since $\mathbb{E}[Z_n \mid U_n] = 0$, we obtain
\[
\mathbb{E}[Z_{n-1}Z_n \mid U_n] = 0.
\]

Taking expectations again,
\[
\mathbb{E}[Z_{n-1}Z_n] = 0.
\]

Thus,
\[
\mathrm{Cov}(Z_{n-1}, Z_n) = 0.
\]

Successive Gibbs samples $Z_{n-1}$ and $Z_n$ are uncorrelated in stationarity. 
However, they are not independent, since both depend on the intermediate variable $U_n$.

### Correlation between successive second moments $Z_{n-1}^2$ and $Z_n^2$}

Recall that
\[
Z_n \mid U_n \sim \mathrm{Unif}\!\big(-a(U_n), a(U_n)\big),
\qquad
a(u) = \sqrt{-2\log u}.
\]

For a symmetric uniform distribution on $(-a,a)$ we have
\[
\mathbb{E}[X^2] = \frac{a^2}{3}.
\]

Therefore,
\[
\mathbb{E}[Z_n^2 \mid U_n]
=
\frac{a(U_n)^2}{3}
=
\frac{-2\log U_n}{3}.
\]

Unlike the first moment case, this conditional second moment depends explicitly on $U_n$.

\medskip

We now compute the covariance using the law of total covariance:
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathbb{E}\!\left[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2 \mid U_n)
\right]
+
\mathrm{Cov}\!\left(
\mathbb{E}[Z_{n-1}^2 \mid U_n],
\mathbb{E}[Z_n^2 \mid U_n]
\right).
\]

In the Gibbs sampler, $Z_n$ is generated from $U_n$ independently of $Z_{n-1}$. Hence, conditional on $U_n$, the variables $Z_{n-1}$ and $Z_n$ are independent, and therefore
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2 \mid U_n) = 0.
\]

Thus,
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Cov}\!\left(
\mathbb{E}[Z_{n-1}^2 \mid U_n],
\mathbb{E}[Z_n^2 \mid U_n]
\right).
\]

From the previous calculation,
\[
\mathbb{E}[Z_{n-1}^2 \mid U_n]
=
\mathbb{E}[Z_n^2 \mid U_n]
=
\frac{-2\log U_n}{3}.
\]

Therefore,
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Var}\!\left(\frac{-2\log U_n}{3}\right).
\]

\medskip

To evaluate this variance, we determine the marginal distribution of $U_n$ in stationarity.  
It can be shown that
\[
T = -2\log U_n \sim \chi^2_3.
\]

Hence,
\[
\mathbb{E}[T] = 3,
\qquad
\mathrm{Var}(T) = 6.
\]

It follows that
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Var}(T/3)
=
\frac{6}{9}
=
\frac{2}{3}.
\]

Successive second moments are positively correlated:
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2) = \frac{2}{3} \neq 0.
\]
Thus, although successive Gibbs samples $Z_{n-1}$ and $Z_n$ are uncorrelated, their squared values exhibit positive correlation due to the shared dependence on the intermediate variable $U_n$.


# Problem 1
## a)

```{r}
load(url("https://www.math.ntnu.no/emner/TMA4300/2026v/random-intercept-data.RData"))

#str(data)
#head(data)
```


```{r}
mod <- glmmTMB(y ~ (1|i), data=data)
summary(mod)
```

The maximum likelihood estimate of the overall mean is $\hat{\mu} = 10.1888$. The estimated variance of the random intercepts is $\widehat{\mathrm{Var}}(\gamma_i) = 0.2301$, corresponding to $1/\nu$, and the estimated residual variance is $\widehat{\mathrm{Var}}(\epsilon_{ij}) = 0.7560$, corresponding to $1/\tau$. These estimates will later be used to check that the Gibbs sampler implementation produces reasonable results.

## b) Full conditional distributions

From the hierarchical model

\[
y_{ij} \mid \mu,\gamma_i,\tau \sim N(\mu+\gamma_i,\tau^{-1}),
\]

\[
\gamma_i \mid \nu \sim N(0,\nu^{-1}),
\]

with priors

\[
\pi(\mu)\propto 1, \qquad \tau \sim \text{Gamma}(\alpha,\alpha), \qquad \nu \sim \text{Gamma}(\alpha,\alpha),
\]

we derive the following full conditional distributions.

Let \(N=\sum_{i=1}^{m} n_i\).


#### Full conditional of \( \mu \)

Combining the likelihood with the flat prior for \(\mu\), we obtain

\[
\mu \mid \gamma,\tau,y \sim N(\bar{y}_\gamma,(N\tau)^{-1}),
\]

where

\[
\bar{y}_\gamma
=
\frac{1}{N}
\sum_{i=1}^{m}\sum_{j=1}^{n_i}(y_{ij}-\gamma_i).
\]


#### Full conditional of \( \gamma_i \)

Using the likelihood contribution from cluster \(i\) together with the prior \(\gamma_i \sim N(0,\nu^{-1})\), we obtain

\[
\gamma_i \mid \mu,\tau,\nu,y
\sim
N(m_i,(\nu+n_i\tau)^{-1}),
\]

where

\[
m_i
=
\frac{\tau\sum_{j=1}^{n_i}(y_{ij}-\mu)}{\nu+n_i\tau}.
\]


#### Full conditional of \( \tau \)

Let

\[
S=\sum_{i=1}^{m}\sum_{j=1}^{n_i}(y_{ij}-\mu-\gamma_i)^2.
\]

Combining the Gaussian likelihood with the prior \(\tau \sim \text{Gamma}(\alpha,\alpha)\) gives

\[
\tau \mid \mu,\gamma,\nu,y
\sim
\text{Gamma}\left(\alpha+\frac{N}{2},\,\alpha+\frac{S}{2}\right).
\]


#### Full conditional of \( \nu \)

Using the prior for \(\gamma_i\) and \(\nu \sim \text{Gamma}(\alpha,\alpha)\), we obtain

\[
\nu \mid \mu,\gamma,\tau,y
\sim
\text{Gamma}\left(\alpha+\frac{m}{2},\,\alpha+\frac{1}{2}\sum_{i=1}^{m}\gamma_i^2\right).
\]

## c) Implementation of a Gibbs sampler

```{r}
gibbs_sampler <- function(data, n_iter = 20000, alpha = 0.0001){

y <- data$y
i <- data$i

groups <- unique(i)
m <- length(groups)
N <- length(y)

n_i <- table(i)

mu <- mean(y)
gamma <- rep(0,m)
tau <- 1
nu <- 1

samples <- matrix(NA, n_iter, m+3)

for (iter in 1:n_iter){

# update mu
y_gamma <- y - gamma[i]
mu <- rnorm(1, mean(y_gamma), sqrt(1/(N*tau)))

# update gamma_i
for (k in 1:m){

y_k <- y[i==groups[k]]
n_k <- length(y_k)

var_gamma <- 1/(nu + n_k*tau)
mean_gamma <- (tau*sum(y_k-mu))/(nu+n_k*tau)

gamma[k] <- rnorm(1, mean_gamma, sqrt(var_gamma))
}

# update tau
res <- y - mu - gamma[i]
tau <- rgamma(1, alpha + N/2, alpha + sum(res^2)/2)

# update nu
nu <- rgamma(1, alpha + m/2, alpha + sum(gamma^2)/2)

samples[iter,] <- c(mu, gamma, 1/tau, 1/nu)
}

return(samples)
}
```

```{r}
samples <- gibbs_sampler(data, 20000)
colMeans(samples)
```

The Gibbs sampler was run for $20000$ iterations to sample from the joint posterior distribution of the model parameters. The posterior mean of the global intercept is approximately $\mu \approx 10.19$, which is very close to the maximum likelihood estimate obtained in part (a). The posterior mean of the residual variance is approximately $1/\tau \approx 0.80$ and the posterior mean of the random intercept variance is approximately $1/\nu \approx 0.25$. These values are close to the corresponding maximum likelihood estimates ($0.756$ and $0.2301$), suggesting that the Gibbs sampler produces reasonable estimates and appears to converge to the correct posterior distribution.

## d)
```{r}
matplot(samples[,c(1,22,23)], type="l")
```

The trace plots for the parameters appear stable and fluctuate around constant levels, indicating that the Gibbs sampler mixes reasonably well and that the Markov chains have reached a stationary distribution. The chain for the global intercept μ is centered around approximately $10.2$, while the residual variance $1/τ$ fluctuates around about $0.8$ and the random intercept variance $1/ν$ around about $0.25$. These values are consistent with the maximum likelihood estimates obtained in part (a). The trace plot for $1/ν$ shows slightly larger variability and occasional spikes compared to the other parameters, which is expected for variance parameters with weakly informative priors, but overall the chain still appears stable.




