---
title: "Project 2 – TMA4300"
author: "Karine Grande & Thea Boge"
date: "2026-03-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1
We suppose that the joint density of $U$, $Z$ up to a normalizing constant is 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]

## a) 
If a joint density is known up to proportionality, the full conditional of one variable is obtained by 
\[
f_{X|Y} (x|y) \propto f^{*}_{U,Z}(u,z) \textit{as a function of x}
\]

### Conditional ditribution of $U|Z=z$

Fix $z$. The joint density becomes 
\[
f^{*}_{U,Z}(u,z) =
\begin{cases}
1, & \text{for } 0 < u < e^{-z^2/2}, \\
0, & \text{elsewhere.}
\end{cases}
\]
As a function of $u$, this is constant on the interval $(0,e^{-x^{\frac{2}{2}}})$. 

The normalizing constant is
\[
\int_0^{e^{-z^2/2}} 1 \, du =e^{-z^2/2}.
\]

Therefore,
\[
f_{U \mid Z}(u \mid z) = \frac{1}{e^{-z^2/2}} = e^{z^2/2},\quad 0 < u < e^{-z^2/2}.
\]

Hence,
\[
U \mid Z = z \sim \mathrm{Unif} \big(0, e^{-z^2/2}\big).
\]. 

### Conditional ditribution of $Z|U=u$
Fix $u$. The support condition is $0 < u < e^{-z^2/2}$.

Solving for $z$:
\[
u < e^{-z^2/2} \quad \Longleftrightarrow \quad \log u < -\frac{z^2}{2}
\]

\[
-2 \log u > z^2
\]

\[
|z| < \sqrt{-2 \log u}
\]

Define
\[
a(u) = \sqrt{-2 \log u}
\]

Therefore, the support of the conditional distribution is
\[
z \in (-a(u), a(u)).
\]

Within this interval the joint density is constant and equal to 1. The normalizing constant is therefore the length of the interval:
\[
\int_{-a(u)}^{a(u)} 1 \, dz = 2a(u).
\]

Thus, the full conditional density of \( Z \mid U = u \) is

\[
f_{Z \mid U}(z \mid u) = \frac{1}{2a(u)},\quad |z| < a(u),
\]

where
\[
a(u) = \sqrt{-2 \log u}.
\]

Thus,
\[
Z \mid U = u \sim \mathrm{Unif}\big(-a(u), a(u)\big).
\]

### Gibbs Sampler

Given an initial valid point \( (U^{(0)}, Z^{(0)}) \) satisfying
\( 0 < U^{(0)} < e^{-(Z^{(0)})^2/2} \),
iterate for \( t = 1, \dots, n \):

\begin{enumerate}
    \item Sample
    \[
    U^{(t)} \sim 
    \mathrm{Unif}\left(0, e^{-(Z^{(t-1)})^2/2}\right).
    \]

    \item Define
    \[
    a^{(t)} = \sqrt{-2 \log U^{(t)}}.
    \]

    \item Sample
    \[
    Z^{(t)} \sim 
    \mathrm{Unif}\left(-a^{(t)},\, a^{(t)}\right).
    \]
\end{enumerate}

This iterative scheme defines a Markov chain whose stationary distribution is the joint density \( f_{U,Z}(u,z) \).The marginal stationary distribution of \( Z \) is \( \mathcal{N}(0,1) \).

```{r}
gibbs_normal <- function(n, u_init = 0.5, z_init = 0) {
  
  U <- numeric(n)
  Z <- numeric(n)
  
  U[1] <- u_init
  Z[1] <- z_init
  
  for (t in 2:n) {
    
    # Sample U | Z
    U[t] <- runif(1, 0, exp(-(Z[t-1]^2)/2))
    
    # Compute truncation bound
    a <- sqrt(-2 * log(U[t]))
    
    # Sample Z | U
    Z[t] <- runif(1, -a, a)
  }
  
  Z  # return only Z since that is the target
}
```

```{r}
# Run Gibbs sampler
set.seed(123)
Z <- gibbs_normal(10000)

# Burn-in removal
Z_post <- Z[1001:10000]

# Diagnostics
mean(Z_post)
var(Z_post)

hist(Z_post, probability = TRUE, breaks = 40)
curve(dnorm(x), col="red", add=TRUE)
```
The posterior sample mean is approximately
\[
\hat{\mu} \approx -0.019,
\]
and the empirical variance is
\[
\hat{\sigma}^2 \approx 0.97.
\]

Both values are close to the theoretical mean and variance,
\[
\mathbb{E}[Z] = 0,\qquad \mathrm{Var}(Z) = 1,
\]
of a standard normal distribution.

The histogram aligns well with the standard normal density curve, confirming that the Gibbs sampler correctly targets
\[
Z \sim \mathcal{N}(0,1).
\]

## b)
We determine the marginal distribution of \( Z \) by integrating out \( U \).

Recall that the joint density (up to normalization) is

\[
f^*_{U,Z}(u,z) =
\begin{cases}
1, & 0 < u < e^{-z^2/2}, \\
0, & \text{otherwise}.
\end{cases}
\]

The marginal density of \( Z \) is therefore

\[
f_Z(z)
=
\int_0^{e^{-z^2/2}} 1 \, du
=
e^{-z^2/2}.
\]

Thus,

\[
f_Z(z) \propto e^{-z^2/2}.
\]


Since
\[
\int_{-\infty}^{\infty} e^{-z^2/2} \, dz
=
\sqrt{2\pi},
\]
the normalized density is

\[
f_Z(z)
=
\frac{1}{\sqrt{2\pi}} e^{-z^2/2}.
\]

Therefore,
\[
Z \sim \mathcal{N}(0,1).
\]
Hence, discarding the \( U \)-values from the Gibbs sampler yields a sample from a standard normal distribution.

## c)
### Correlation between successive Gibbs samples $Z_{n-1}$ and $Z_n$}

From the Gibbs sampler we have
\[
Z_n \mid U_n \sim \mathrm{Unif}\!\big(-a(U_n), a(U_n)\big),
\qquad
a(u) = \sqrt{-2 \log u}.
\]

This conditional distribution is symmetric around zero. Therefore,
\[
\mathbb{E}[Z_n \mid U_n] = 0.
\]

By the law of total expectation,
\[
\mathbb{E}[Z_n]
=
\mathbb{E}\!\left[\mathbb{E}[Z_n \mid U_n]\right]
=
\mathbb{E}[0]
=
0.
\]

Since the Gibbs sampler has stationary distribution $Z_n \sim \mathcal{N}(0,1)$, it follows that
\[
\mathbb{E}[Z_{n-1}] = \mathbb{E}[Z_n] = 0.
\]

We now compute the covariance. By definition,
\[
\mathrm{Cov}(Z_{n-1}, Z_n)
=
\mathbb{E}[Z_{n-1}Z_n]
-
\mathbb{E}[Z_{n-1}]\mathbb{E}[Z_n].
\]

Because both expectations are zero, this reduces to
\[
\mathrm{Cov}(Z_{n-1}, Z_n)
=
\mathbb{E}[Z_{n-1}Z_n].
\]

Using the tower property (law of iterated expectation),
\[
\mathbb{E}[Z_{n-1}Z_n]
=
\mathbb{E}\!\left[
\mathbb{E}[Z_{n-1}Z_n \mid U_n]
\right].
\]

In the Gibbs sampler, $Z_n$ is generated from $U_n$ independently of $Z_{n-1}$. Hence, conditional on $U_n$, the variables $Z_{n-1}$ and $Z_n$ are independent. Therefore,
\[
\mathbb{E}[Z_{n-1}Z_n \mid U_n]
=
Z_{n-1} \, \mathbb{E}[Z_n \mid U_n].
\]

Since $\mathbb{E}[Z_n \mid U_n] = 0$, we obtain
\[
\mathbb{E}[Z_{n-1}Z_n \mid U_n] = 0.
\]

Taking expectations again,
\[
\mathbb{E}[Z_{n-1}Z_n] = 0.
\]

Thus,
\[
\mathrm{Cov}(Z_{n-1}, Z_n) = 0.
\]

Successive Gibbs samples $Z_{n-1}$ and $Z_n$ are uncorrelated in stationarity. 
However, they are not independent, since both depend on the intermediate variable $U_n$.

### Correlation between successive second moments $Z_{n-1}^2$ and $Z_n^2$}

Recall that
\[
Z_n \mid U_n \sim \mathrm{Unif}\!\big(-a(U_n), a(U_n)\big),
\qquad
a(u) = \sqrt{-2\log u}.
\]

For a symmetric uniform distribution on $(-a,a)$ we have
\[
\mathbb{E}[X^2] = \frac{a^2}{3}.
\]

Therefore,
\[
\mathbb{E}[Z_n^2 \mid U_n]
=
\frac{a(U_n)^2}{3}
=
\frac{-2\log U_n}{3}.
\]

Unlike the first moment case, this conditional second moment depends explicitly on $U_n$.

\medskip

We now compute the covariance using the law of total covariance:
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathbb{E}\!\left[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2 \mid U_n)
\right]
+
\mathrm{Cov}\!\left(
\mathbb{E}[Z_{n-1}^2 \mid U_n],
\mathbb{E}[Z_n^2 \mid U_n]
\right).
\]

In the Gibbs sampler, $Z_n$ is generated from $U_n$ independently of $Z_{n-1}$. Hence, conditional on $U_n$, the variables $Z_{n-1}$ and $Z_n$ are independent, and therefore
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2 \mid U_n) = 0.
\]

Thus,
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Cov}\!\left(
\mathbb{E}[Z_{n-1}^2 \mid U_n],
\mathbb{E}[Z_n^2 \mid U_n]
\right).
\]

From the previous calculation,
\[
\mathbb{E}[Z_{n-1}^2 \mid U_n]
=
\mathbb{E}[Z_n^2 \mid U_n]
=
\frac{-2\log U_n}{3}.
\]

Therefore,
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Var}\!\left(\frac{-2\log U_n}{3}\right).
\]

\medskip

To evaluate this variance, we determine the marginal distribution of $U_n$ in stationarity.  
It can be shown that
\[
T = -2\log U_n \sim \chi^2_3.
\]

Hence,
\[
\mathbb{E}[T] = 3,
\qquad
\mathrm{Var}(T) = 6.
\]

It follows that
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2)
=
\mathrm{Var}(T/3)
=
\frac{6}{9}
=
\frac{2}{3}.
\]

Successive second moments are positively correlated:
\[
\mathrm{Cov}(Z_{n-1}^2, Z_n^2) = \frac{2}{3} \neq 0.
\]
Thus, although successive Gibbs samples $Z_{n-1}$ and $Z_n$ are uncorrelated, their squared values exhibit positive correlation due to the shared dependence on the intermediate variable $U_n$.
